Precise Data Flow Analysis\\in the Presence of Correlated Method Calls

### 2014-12-11 content/background.tex
minor changes and comments in section 3

\section{Background}\label{sec:bg}
%The purpose of the correlated-calls analysis is to solve IFDS problems more precisely than using the standard IFDS algorithm by ruling out some infeasible paths. The correlated-calls analysis works by transforming an IFDS problem to an IDE problem, solving the IDE problem, and transforming the IDE result to a solution to the original IFDS problem. 
This section defines basic terminology and presents the IFDS and IDE algorithms.

@@

A call site is \textit{monomorphic} if it always calls the same procedure. In an object-oriented language, 
a call site $r.m(\ldots)$ can dynamically dispatch to multiple methods depending on the runtime
type of the object pointed to by the receiver $r$.
A call site that calls multiple procedures is called
\textit{polymorphic}.
We define a function $\textsf{lookup}$ to specify the dynamic dispatch: 
if $s$ is the signature of $m$ and $t$ is the runtime type of the object
pointed to by $r$, $\textsf{lookup}(s,t)$ gives the procedure that will
be invoked by the call $r.m(\ldots)$. We also define a kind of inverse $\tau$
of $\textsf{lookup}$: given a signature $s$ and a specific invoked procedure
$f$, $\tau(s,f)$ gives the set of all runtime types of $r$ that cause $r.m(\ldots)$
to dispatch to $f$: $\tau(s,f) = \{t\mid \textsf{lookup}(s,t)=f\}$.

@@

A path through the control-flow supergraph is \textit{valid} if every end-to-return edge on the path returns
to the site of the most recent unmatched call. The set of all valid paths from the program entry point to
a node $n$ is denoted $\ivp(n)$.
\otodo{Should we give a formal definition?}

@@

\otodo{The definitions that Marianna had were for \emph{complete}
lattices/semilattices. All of the lattices in the context of IFDS/IDE
are/must be complete. I'll just say that here.}
A \textit{complete lattice} is a partially ordered set $(S,\sqsubseteq)$ in which every subset has a least upper bound and a greatest lower bound.
A \textit{complete meet semilattice} is a partially ordered set in which every subset need only have a greatest lower bound.
The symbols $\bot$ and $\top$ are used to denote the greatest lower bound of $S$ and of the empty set, respectively.
Since all of the (semi)lattices discussed in this paper are required to be complete, we will henceforth leave out the
\textit{complete} qualifier.

@@

The IFDS algorithm is notable because it computes a meet over valid paths solution in polynomial time.
Most other interprocedural analysis algorithms are either imprecise due to invalid paths,
general but do not run in polynomial time~\cite{knoop1992interprocedural,pnueli1981two}, or handle a very specific set of problems~\cite{knoop1993efficient}.

@@

The input to the IFDS algorithm is specified as
%Formally, a data-flow analysis problem is suitable for an IFDS analysis if it can be encoded as an IFDS problem
  $(G^*,\,D,\,F,\,M_F,\,\sqcap)$,
where $G^*=(N^*,\,E^*)$ is the supergraph of the input program with nodes $N^*$ and edges $E^*$,
$D$ is a finite set of \textit{data-flow facts},
$F$ is a set of distributive dataflow functions of type $2^D\to2^D$,
$M_F:\,E^*\to F$ assigns a dataflow function to each supergraph edge,
and $\sqcap$ is the \textit{meet operator} on the powerset $2^D$, either union or intersection.
In our presentation, the meet will always be union, but all of the results apply dually when the
meet is intersection.

@@

The representation relation decomposes a flow function into functions
(edges) that operate on each fact individually. This is possible due
to distributivity: applying the flow function to a set of facts is
equivalent to applying it on each fact individually and
taking the union of the results.

@@

It is this graph of representation relations that the IFDS algorithm operates on.
In this graph, called the \textit{exploded supergraph}, each node is a pair $(n,d)$, where
$n\in N^*$ is a node of the control-flow supergraph and $d$ is an element of $D\cup\{0\}$.
For each edge $(n \to n') \in E^*$, the exploded supergraph contains a set of edges
$(n,d_i) \to (n',d_j)$, which form the representation relation of the dataflow function
$M_F(n \to n')$. The IFDS algorithm finds all exploded supergraph edges that are reachable by
\textit{realizable} paths in the exploded supergraph. A path
is \textit{realizable} if its projection to the (non-exploded) supergraph is a valid
path (i.e. if it is of the form $(n_0, d_0) \to (n_1,d_1) \to \cdots \to (n_m,d_m)$
and $n_0 \to n_1 \to \cdots \to n_m$ is a valid path).

@@

\subsubsection{Overview of the IDE Algorithm}\label{sec:ideoverview}
Just as any distributive function from $2^D$ to $2^D$ can be represented with a
representation relation, it is also possible to represent any distributive functions from
$\textsf{Env}(D,L)$
to
$\textsf{Env}(D,L)$
with a \textit{pointwise representation}. A pointwise representation is a bipartite graph
with the same nodes~\footnote{The IDE literature uses the symbol $\Lambda$ for the node that
    is denoted $\mathbf0$ in the IFDS literature.}
and edges as a representation relation, except that each edge is labelled
with a \textit{micro-function}, which is a function from $L$ to $L$. 
\commentout{
The pointwise representation of an environment transformer 
$t : \textsf{Env}(D,L) \to \textsf{Env}(D,L)$
is defined as
\newcommand{\microedge}[3]{#1\xrightarrow{#3}#2}
\begin{align*}
    R_t =& \{ \microedge{\Lambda}{\Lambda}{\lambda l.l} \} \cup
    \{\microedge{\Lambda}{d_j}{\lambda l.t(\Omega)(d_j)} \mid d_j \in D\}\\
    &\cup
    \{\microedge{d_i}{d_j}{\lambda l.t(\Omega[d_i\to l])(d_j)} \mid d_i, d_j \in D\}.
\end{align*}
}
Let $\Omega = \lambda d.\top$ be the environment that maps every element of $D$ to $\top$.
Thanks to distributivity, every environment transformer 
$t : \textsf{Env}(D,L) \to \textsf{Env}(D,L)$
can be decomposed into its effect on $\Omega$ and on a set of environments $\Omega[d_i\to l]$
that map every element except one ($d_i$) to $\top$:
\[
    t(m)(d_j) = \lambda l. t(\Omega)(d_j) \sqcap \bigsqcap_{d_i\in D} \lambda l. t(\Omega[d_i\to l])(d_j).
\]
The functions $\lambda l. \cdots$ in this decomposition are the micro-functions that
appear on the edges of the pointwise representation edges from $\Lambda$ to each $d_j$ and from
each $d_i$ to each $d_j$.%
\footnote{The IDE paper defines a more complicated but equivalent set of micro-functions
that eliminate some duplication of computation.}
The absence of an edge in the pointwise representation from some $d_i$ to some $d_j$ is
equivalent to an edge with micro-function $\lambda l.\top$.

@@

Similar to the IFDS algorithm, a practical implementation of the IDE algorithm requires
the input dataflow functions to be provided in their pointwise representation
as exploded supergraph edges labelled with micro-functions.
\mtodo{Should we cite the ``Practical extensions'' paper here that talks about the on-demand IFDS algorithm?}
Specifically,
the input is generally provided as a function $f: N^* \times D \times N^* \to (D \to F)$,
where $F$ is the set of representations of micro-functions from $L$ to $L$.
Given a (non-exploded) supergraph edge $n \to n'$ and a dataflow fact $d$,
$f(n,d,n')$ returns a map that gives for each fact $d'$ the micro-function $f$
that appears on the exploded supergraph edge $(n,d) \to (n',d')$.
Like in the IFDS algorithm, the function $f$ can be split up into separate functions that handle the cases when
the $n\to n'$ is an intraprocedural edge, a call-to-start edge, or an end-to-return edge.

@@

>>>>

\section{Background}\label{sec:bg}
%The purpose of the correlated-calls analysis is to solve IFDS problems more precisely than using the standard IFDS algorithm by ruling out some infeasible paths. The correlated-calls analysis works by transforming an IFDS problem to an IDE problem, solving the IDE problem, and transforming the IDE result to a solution to the original IFDS problem. 
This section defines basic terminology and presents the IFDS and IDE algorithms.
\ftodo{
  Shall we add a sentence saying that readers familiar with these algorithms may skip this section?
}

@@

A call site is \textit{monomorphic} if it always calls the same procedure. In an object-oriented language, 
a call site $r.m(\ldots)$ can dynamically dispatch to multiple methods depending on the runtime
type of the object pointed to by the receiver $r$.
A call site that calls multiple procedures is called
\textit{polymorphic}.
We define a function $\textsf{lookup}$ to specify the dynamic dispatch: 
if $s$ is the signature of $m$ and $t$ is the runtime type of the object
pointed to by $r$, $\textsf{lookup}(s,t)$ gives the procedure that will
be invoked by the call $r.m(\ldots)$. We also define a function $\tau$
that may be viewed as the inverse
of $\textsf{lookup}$: given a signature $s$ and a specific invoked procedure
$f$, $\tau(s,f)$ gives the set of all runtime types of $r$ that cause $r.m(\ldots)$
to dispatch to $f$: $\tau(s,f) = \{t\mid \textsf{lookup}(s,t)=f\}$.

@@

A path through the control-flow supergraph is \textit{valid} if every end-to-return edge on the path returns
to the site of the most recent unmatched call. 
\ftodo{Can we explain/define the notion of the ``most recent unmatched call'' more precisely?}
The set of all valid paths from the program entry point to
a node $n$ is denoted $\ivp(n)$.
\otodo{Should we give a formal definition?}

@@

\otodo{The definitions that Marianna had were for \emph{complete}
lattices/semilattices. All of the lattices in the context of IFDS/IDE
are/must be complete. I'll just say that here.}
\ftodo{Would it make sense to first the define the notion of a lattice without referring to its completeness?
       This could be followed by defining what it means for a (semi)lattice to be complete, and then we
       could say that all lattices considered in this paper are complete, etc.}
\ftodo{Can we define the symbols $\sqcup$ and $\sqcap$ for join and meet here?}
A \textit{complete lattice} is a partially ordered set $(S,\sqsubseteq)$ in which every subset has a least upper bound and a greatest lower bound.
A \textit{complete meet semilattice} is a partially ordered set in which every subset need only have a greatest lower bound.
The symbols $\bot$ and $\top$ are used to denote the greatest lower bound of $S$ and of the empty set, respectively.
Since all of the (semi)lattices discussed in this paper are required to be complete, we will henceforth leave out the
\textit{complete} qualifier.

@@

The IFDS algorithm is notable because it computes a meet-over-valid paths solution in polynomial time.
Most other interprocedural analysis algorithms are either: 
  (i) imprecise due to invalid paths,
  (ii) general but do not run in polynomial time~\cite{knoop1992interprocedural,pnueli1981two}, or 
  (iii) handle a very specific set of problems~\cite{knoop1993efficient}.

@@

The input to the IFDS algorithm is specified as
%Formally, a data-flow analysis problem is suitable for an IFDS analysis if it can be encoded as an IFDS problem
  $(G^*,\,D,\,F,\,M_F,\,\sqcap)$,
where $G^*=(N^*,\,E^*)$ is the supergraph of the input program with nodes $N^*$ and edges $E^*$,
$D$ is a finite set of \textit{data-flow facts},
$F$ is a set of distributive dataflow functions of type $2^D\to2^D$,
$M_F:\,E^*\to F$ assigns a dataflow function to each supergraph edge,
and $\sqcap$ is the \textit{meet operator} on the powerset $2^D$, either union or intersection.
In our presentation, the meet operator will always be union, but all of the results apply dually when the
meet is intersection.

@@

The representation relation decomposes a flow function into functions
(edges) that operate on each fact individually. This is possible due
to distributivity: applying the flow function to a set of facts is
equivalent to applying it on each fact individually and then
taking the union of the results.

@@

It is this graph of representation relations that the IFDS algorithm operates on.
In this graph, called the \textit{exploded supergraph}, each node is a pair $(n,d)$, where
$n\in N^*$ is a node of the control-flow supergraph and $d$ is an element of $D\cup\{0\}$.
For each edge $(n \to n') \in E^*$, the exploded supergraph contains a set of edges
$(n,d_i) \to (n',d_j)$, which form the representation relation of the dataflow function
$M_F(n \to n')$. The IFDS algorithm finds all exploded supergraph edges that are reachable by
\textit{realizable} paths in the exploded supergraph. A path
is \textit{realizable} if its projection to the (non-exploded) supergraph is a valid
path (i.e., if it is of the form $(n_0, d_0) \to (n_1,d_1) \to \cdots \to (n_m,d_m)$
and where $n_0 \to n_1 \to \cdots \to n_m$ is a valid path).

@@

\subsubsection{Overview of the IDE Algorithm}\label{sec:ideoverview}
Just as any distributive function from $2^D$ to $2^D$ can be represented with a
representation relation, it is also possible to represent any distributive functions from
$\textsf{Env}(D,L)$
to
$\textsf{Env}(D,L)$
with a \textit{pointwise representation}. A pointwise representation is a bipartite graph
with the same nodes~\footnote{The IDE literature uses the symbol $\Lambda$ for the node that
    is denoted $\mathbf0$ in the IFDS literature.}
and edges as a representation relation, except that each edge is labelled
with a \textit{micro-function}, which is a function from $L$ to $L$. 
\commentout{
The pointwise representation of an environment transformer 
$t : \textsf{Env}(D,L) \to \textsf{Env}(D,L)$
is defined as
\newcommand{\microedge}[3]{#1\xrightarrow{#3}#2}
\begin{align*}
    R_t =& \{ \microedge{\Lambda}{\Lambda}{\lambda l.l} \} \cup
    \{\microedge{\Lambda}{d_j}{\lambda l.t(\Omega)(d_j)} \mid d_j \in D\}\\
    &\cup
    \{\microedge{d_i}{d_j}{\lambda l.t(\Omega[d_i\to l])(d_j)} \mid d_i, d_j \in D\}.
\end{align*}
}
Let $\Omega = \lambda d.\top$ be the environment that maps every element of $D$ to $\top$.
\ftodo{
   What is the relationship between $\Omega$ and $\Delta$?
}
Thanks to distributivity, every environment transformer 
$t : \textsf{Env}(D,L) \to \textsf{Env}(D,L)$
can be decomposed into its effect on $\Omega$ and on a set of environments $\Omega[d_i\to l]$
that map every element except one ($d_i$) to $\top$:
\[
    t(m)(d_j) = \lambda l. t(\Omega)(d_j) \sqcap \bigsqcap_{d_i\in D} \lambda l. t(\Omega[d_i\to l])(d_j).
\]
The functions $\lambda l. \cdots$ in this decomposition are the micro-functions that
appear on the edges of the pointwise representation edges from $\Lambda$ to each $d_j$ and from
each $d_i$ to each $d_j$.%
\footnote{The IDE paper defines a more complicated but equivalent set of micro-functions
that eliminate some duplication of computation.}
The absence of an edge in the pointwise representation from some $d_i$ to some $d_j$ is
equivalent to an edge with micro-function $\lambda l.\top$.

@@

Similar to the IFDS algorithm, a practical implementation of the IDE algorithm requires
the input dataflow functions to be provided in their pointwise representation
as exploded supergraph edges labelled with micro-functions.
\mtodo{Should we cite the ``Practical extensions'' paper here that talks about the on-demand IFDS algorithm?}
Specifically,
the input is generally provided as a function $f: N^* \times D \times N^* \to (D \to F)$,
where $F$ is the set of representations of micro-functions from $L$ to $L$.
Given a (non-exploded) supergraph edge $n \to n'$ and a dataflow fact $d$,
$f(n,d,n')$ returns a map that gives, for each fact $d'$, the micro-function $f$
that appears on the exploded supergraph edge $(n,d) \to (n',d')$.
Like in the IFDS algorithm, the function $f$ can be split up into separate functions that handle the cases when
the $n\to n'$ is an intraprocedural edge, a call-to-start edge, or an end-to-return edge.

@@

\ftodo{Would it make sense to discuss an example of an IDE problem here?  Perhaps we can do this
  without using much space by referring to an existing figure elsewhere in the paper?}
